Structuring College Access: The Market Segment Model and College Board Geo-markets

### 2024-10-18 soc_of_ed/soc_of_ed_geomarket.qmd
revising front end

---
title: "Structuring College Access: The Market Segment Model and College Board Geo-markets"
#author: ["Ozan Jaquette (UCLA)", "Karina Salazar (University of Arizona)"]
bibliography: ./../assets/bib/cb_geomarket_bib.bib
csl: ./../assets/bib/apa.csl
format:
  pdf:
    documentclass: article
    number-sections: true
    number-depth: 2
    #extensions: [andrewheiss/wordcount]
    fig_caption: true
    keep_tex: true  # Optional: Keeps the intermediate .tex file for debugging
editor: visual
geometry: "margin=1in"
fontsize: 11pt
header-includes:
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhf{}
  - \cfoot{\thepage}
  - \fancypagestyle{plain}{
      \fancyhf{}
      \cfoot{\thepage}
    }
  - \usepackage{booktabs}
  - \usepackage{multirow}
  - \usepackage{array}
  - \usepackage{colortbl}
  - \usepackage{longtable}
  - \usepackage{caption}
  - \usepackage{pdflscape} 
  - \usepackage{rotating}
  - \usepackage{float}       # Required for [ht] specifier
  - \usepackage{graphicx}    # Required for \resizebox  
  - |
    \captionsetup{
      justification=raggedright,
      singlelinecheck=false
    }
  - \setlength{\aboverulesep}{0pt}
  - \setlength{\belowrulesep}{0pt}
  - \setlength{\extrarowheight}{0pt}
  - \setlength{\cmidrulekern}{0pt}
  - \renewcommand{\arraystretch}{1.0}
---

@@

Geomarkets are geographic borders that disaggregate states and large metropolitan areas into "community-based enrollment markets" [@RN4982, p. 14] for recruiting purposes. @fig-nyc-geomarkets shows New York City area Geomarkets. Geomarkets are the basis for the College Board Enrollment Planning Service (EPS), founded in 1984 and still active today. EPS software recommends which Geomarkets a college should recruit from and which schools/communities they should prioritize within targeted Geomarkets.^[In 1995, 37% of 4-year public institutions and 49% of 4-year private institutions used EPS [@RN5021].]

Geomarkets were incorporated into the _College Board_ "student list" product named Student Search Service. Student lists have been the primary source of lead generation in U.S. higher education for over 50 years [@belkin2019-studata; @list_biz].^[@RN4895 reported that 87% of private and 86% of public four-year institutions purchase student lists.] Lists contain contact information for prospective students. The Student Search Service database consists of College Board test-takers. Colleges pay a fee for each prospect (e.g., \$0.50). Colleges control which prospect profiles they purchase by selecting search filters, such as high school graduation year, SAT score, AP score, state, etc. Geomarket filters enable colleges to include/exclude prospects from particular Geomarkets. An analysis of 830 student lists purchased by 14 public universities found that 16% of purchases filtered on Geomarket [@RN5012].

@@

1. What is the socioeconomic and racial variation between Geomarkets in metropolitan areas? How does this variation change over time? 
1. How does the socioeconomic and racial composition of included versus excluded prospects vary when student list purchases filter on particular Geomarkets?

@@

**DATA SOURCES**. To answer RQ1, we utilize census tract-level data from the 1980 Decennial Census, 2000 Decennial Census, and 2020 5-year American Community Survey (ACS). Variables of interest include: race/ethnicity; mean and median household income; percent of households below poverty line; and BA attainment of people aged 25+. We obtained shapefiles for College Board Geomarkets from a 2012 *R-bloggers* post. We assigned census tracts to Geomarkets by implementing a partial spatial join.^[Census tracts that intersected multiple Geomarkets were partially assigned based on the proportion of land area. Currently, our spatial join for the 2020 ACS is causing some invalid geometries that we will fix prior to AEFP.]

To answer RQ2, we issued public records requests about student lists purchased by 14 public universities from 2016 through 2020. For 414 College Board student list purchases, associated with about 2.6 million prospects, we received both (1) the order summary -- showing which search filters were utilized -- and (2) the de-identified prospect-level data, including race, ethnicity, zip code, and high school. @fig-asu-order-448922 shows the partial order summary, which we converted to tabular data, for a purchase by Arizona State University that filtered on the 2020 high school graduating class, PSAT scores of 1070 to 1180, and selected Geomarkets.^[Analyses for RQ2 additionally utilize data on public and private high schools from NCES.]

@@

\newpage
```{r fig-asu-order-448922, echo=FALSE, fig.cap="Partial order summary from a College Board student list purchase", fig.width=10, fig.height=7, out.extra='angle=90', out.width='110%'}
knitr::include_graphics(path = file.path('.','..','results','figures','asu_order_448922_Redacted.pdf'))
```

\newpage
```{r fig-socal-hispanic, echo=FALSE, fig.cap="Screenshot of Los Angeles area Geomarkets, showing 2020 percent Hispanic by census tract", fig.width=10, fig.height=7, out.extra='angle=90', out.width='110%'}
knitr::include_graphics(path = file.path('.','..','results','figures','socal_hispanic.png'))
```

>>>>

---
title: "Structuring College Access: The Market Segment Model and College Board Geo-markets"
#author: ["Ozan Jaquette (UCLA)", "Karina Salazar (University of Arizona)"]
bibliography: ./../assets/bib/cb_geomarket_bib.bib
csl: ./../assets/bib/apa.csl
format:
  pdf:
    documentclass: article
    number-sections: true
    number-depth: 2
    #extensions: [andrewheiss/wordcount]
    fig_caption: true
    keep_tex: true  # Optional: Keeps the intermediate .tex file for debugging
editor: visual
geometry: "margin=1in"
fontsize: 11pt
header-includes:
  - \usepackage{setspace}
  #- \doublespacing
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhf{}
  - \cfoot{\thepage}
  - \fancypagestyle{plain}{
      \fancyhf{}
      \cfoot{\thepage}
    }
  - \usepackage{booktabs}
  - \usepackage{multirow}
  - \usepackage{array}
  - \usepackage{colortbl}
  - \usepackage{longtable}
  - \usepackage{caption}
  - \usepackage{pdflscape} 
  - \usepackage{rotating}
  - \usepackage{float}       # Required for [ht] specifier
  - \usepackage{graphicx}    # Required for \resizebox  
  - |
    \captionsetup{
      justification=raggedright,
      singlelinecheck=false
    }
  - \setlength{\aboverulesep}{0pt}
  - \setlength{\belowrulesep}{0pt}
  - \setlength{\extrarowheight}{0pt}
  - \setlength{\cmidrulekern}{0pt}
  - \renewcommand{\arraystretch}{1.0}
---

@@

Geomarkets are geographic borders that disaggregate states and large metropolitan areas into "community-based enrollment markets" [@RN4982, p. 14] for recruiting purposes. @fig-nyc-geomarkets shows New York City area Geomarkets. Geomarkets are the basis for the College Board Enrollment Planning Service (EPS), founded in 1984 and still active today. EPS software recommends which Geomarkets a college should recruit from and which schools/communities they should prioritize within targeted Geomarkets.[^1]

[^1]: In 1995, 37% of 4-year public institutions and 49% of 4-year private institutions used EPS [@RN5021].

Geomarkets were incorporated into the *College Board* "student list" product named Student Search Service. Student lists have been the primary source of lead generation in U.S. higher education for over 50 years [@belkin2019-studata; @list_biz].[^2] Lists contain contact information for prospective students. The Student Search Service database consists of College Board test-takers. Colleges pay a fee for each prospect (e.g., \$0.50). Colleges control which prospect profiles they purchase by selecting search filters, such as high school graduation year, SAT score, AP score, state, etc. Geomarket filters enable colleges to include/exclude prospects from particular Geomarkets. An analysis of 830 student lists purchased by 14 public universities found that 16% of purchases filtered on Geomarket [@RN5012].

[^2]: @RN4895 reported that 87% of private and 86% of public four-year institutions purchase student lists.

@@

1.  What is the socioeconomic and racial variation between Geomarkets in metropolitan areas? How does this variation change over time?
2.  How does the socioeconomic and racial composition of included versus excluded prospects vary when student list purchases filter on particular Geomarkets?

@@

**DATA SOURCES**. To answer RQ1, we utilize census tract-level data from the 1980 Decennial Census, 2000 Decennial Census, and 2020 5-year American Community Survey (ACS). Variables of interest include: race/ethnicity; mean and median household income; percent of households below poverty line; and BA attainment of people aged 25+. We obtained shapefiles for College Board Geomarkets from a 2012 *R-bloggers* post. We assigned census tracts to Geomarkets by implementing a partial spatial join.[^3]

[^3]: Census tracts that intersected multiple Geomarkets were partially assigned based on the proportion of land area. Currently, our spatial join for the 2020 ACS is causing some invalid geometries that we will fix prior to AEFP.

To answer RQ2, we issued public records requests about student lists purchased by 14 public universities from 2016 through 2020. For 414 College Board student list purchases, associated with about 2.6 million prospects, we received both (1) the order summary -- showing which search filters were utilized -- and (2) the de-identified prospect-level data, including race, ethnicity, zip code, and high school. @fig-asu-order-448922 shows the partial order summary, which we converted to tabular data, for a purchase by Arizona State University that filtered on the 2020 high school graduating class, PSAT scores of 1070 to 1180, and selected Geomarkets.[^4]

[^4]: Analyses for RQ2 additionally utilize data on public and private high schools from NCES.

@@

```{r fig-asu-order-448922, echo=FALSE, fig.cap="Partial order summary from a College Board student list purchase", fig.width=10, fig.height=7, out.extra='angle=90', out.width='110%'}
knitr::include_graphics(path = file.path('.','..','results','figures','asu_order_448922_Redacted.pdf'))
```

```{r fig-socal-hispanic, echo=FALSE, fig.cap="Screenshot of Los Angeles area Geomarkets, showing 2020 percent Hispanic by census tract", fig.width=10, fig.height=7, out.extra='angle=90', out.width='110%'}
knitr::include_graphics(path = file.path('.','..','results','figures','socal_hispanic.png'))
```
